---
title: "Machine learning 1"
author: "Rongyu Chen(A17708249)"
date: "2026-04-23"
output:
  pdf_document: default
  html_document: default
---

#First up kmeans()

Demo od using kmeans() function in base R. Frist make up some data with a known structure.

```{r}
x <- c(rnorm(30,-3),rnorm(30,3))
x <- cbind(x,y = rev(x))
x
plot(x)
```

Now we have some made up data in 'x'. Let's see how kmeans work with this data.

```{r}
k<- kmeans(x,centers = 2, nstart=20)
k
```

> Q. How many points are in each cluster

```{r}
k$size
```

> Q. How we do get to the cluster membership/assignment.

```{r}
k$cluster
```

> Q. How about cluster center?

```{r}
k$centers
```

Now we got to the main results. Let's use them to plot data with the kmeans result

```{r}
plot(x,col=k$cluster)
points(k$centers,col="blue",pch=15, cex=1)
```
K-means clustering is widely used because it is fast and relatively simple: it takes numeric data as input and returns a vector indicating cluster membership. However, a key limitation is that we must specify the number of clusters in advance when using kmeans().

```{r}
k4 <- kmeans(x, centers = 4)
plot(x, col=k4$cluster)
points(k4$centers, pch=15)
```

Scree plot to pick k centers value brute-force
```{r}
k1 <- kmeans(x, centers=1)
k2 <- kmeans(x, centers=2)
k3 <- kmeans(x, centers=3)
k4 <- kmeans(x, centers=4)
k5 <- kmeans(x, centers=5)
```

```{r}
z <- c(k1$tot.withinss,
k2$tot.withinss,
k3$tot.withinss,
k4$tot.withinss)
plot(z, typ="b")
```

```{r}
n <- NULL
for(i in 1:5){
n <- c(n, kmeans(x, centers=i)$tot.withinss)
}
plot(n, type="b")
```


## Now for hclust() - Hierarchical Clustering

We will cluster the same data 'x' with the 'hclust()'. In this case 'hclust()' requires a distance matrix as input.

```{r}
hclust <- hclust(dist(x))
hclust
```

```{r}
plot(hclust)

```

To get our cluster membership victer we need to "cut" the tree with the 'cutree()'

```{r}
grps <- cutree(hclust,h = 8)
grps
```

Now plot the data with the hclust() results.

```{r}
plot(x,col=grps)
```

# Principal Component Analysis (PCA)

## PCA of UK food data

Read data from website and try a few visualizations.

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url,row.names=1)
x
```

```{r}
cols <- rainbow(nrow(x))
barplot(as.matrix(x),col=cols)
```
There is one plot that can be useful for small datasets:
```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

# PCA to the rescue
The main function for performing PCA in base R is prcomp(). Here, I will transpose the data so that the “foods” are arranged in the columns.
```{r}
pca <- prcomp(t(x))
summary(pca)
```
```{r}
cols<- c("orange", "red", "blue", "darkgreen")
plot(pca$x[,1],pca$x[,2], col=cols, pch=19)
```
```{r}
library(ggplot2)
ggplot(pca$x) +
aes(PC1, PC2) +
geom_point(col=cols, size=2)
```
```{r}
pca$rotation
```
PCA looks spuer useful.