---
title: "Class12"
author: "Rongyu Chen(A17708249)"
date: "2026-05-11"
output:
  pdf_document: default
  html_document: default
---

## Section 1. Proportion of G|G in a population

Downloaded a CSV file from Ensemble < https://uswest.ensembl.org/Homo_sapiens/Variation/Sample?db=cor39975098;v=1s8067378;vdb=variation;vf=88214830#373531_tablePanel >
Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) * 100
```

Now lets look at a different population. I picked GBR

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G
```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr)*100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population
than the MKL population.
Lets now dig into this further.


## Section 4: Population Scale Analysis
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```
```{r}
table(expr$geno)
```

```{r}
median_expression <- tapply(expr$exp, expr$geno, median)
print(median_expression)
```
The sample sizes for each genotype are A/A: 108, A/G: 233, and G/G: 121. The calculated
median expression levels show that A/A is 31.25, A/G is 25.06, and G/G is 20.07.

```{r}
library(ggplot2)
```


```{r}

ggplot(expr) + aes(x=geno, exp, fill=geno) +
geom_boxplot(notch=TRUE)
```

The A/A genotype exhibits a noticeably higher median expression level compared to the G/G genotype. This suggests that the SNP does affect the expression of ORMDL3, as the homozygous G/G genotype is associated with reduced expression values relative to A/A.
So the SNP affects the expression of ORMDL3. Specifically, the G allele is associated with reduced expression of the gene compared to the A allele.










































