---
title: "Lecture 6 Question 6"
author: "Rongyu Chen(A17708249)"
date: "2026-04-22"
output:
  pdf_document: default
  html_document: default
---

We need a wrapper for bio3d that (1) retrieves one or more PDB files, (2) filters each to a specified chain and atom type, (3) generates a B-factor profile plot for the selected structure, and (4) returns the plotted output.

## First install the package

It provides functions to read PDB files (read.pdb()), subset them (trim.pdb()), and visualize B-factor profiles (plotb3()).

```{r}
install.packages("bio3d", repos = "<https://cloud.r-project.org>") # install.packages("bio3d")

library(bio3d)
```

Generalizing B-factor analysis to work with any collection of PDB structures, allowing B-factors to be plotted for a single selected protein from a given set of PDB IDs.

```{r}
plot_protein <- function(pdb_ids, which = 1, chain = "A", elety = "CA") {
  stopifnot(length(pdb_ids) > 0)  # Require at least one PDB ID
  stopifnot(which >= 1 && which <= length(pdb_ids))  # Validate index
  
  pdbs <- lapply(pdb_ids, read.pdb)  # Load PDB files from RCSB
  
  trims <- lapply(pdbs, function(p) {
    trim.pdb(p, chain = chain, elety = elety)
    # Retain only the specified chain and atom type (e.g., CA)
  })
  
  sel <- trims[[which]]        # Select the desired structure
  b <- sel$atom$b              # Extract B-factor values
  
  plotb3(
    b,                         # Plot B-factors vs residue index
    sse = sel,                 # Add secondary structure annotation
    typ = "l",                 # Line plot
    ylab = "Bfactor",          # Y-axis label
    xlab = "Residue index"     # X-axis label
  )
  
  title(main = paste0(
    "PDB: ", pdb_ids[which],
    " | chain=", chain,
    " | elety=", elety
  ))
  
  invisible(list(
    selected_pdb = pdb_ids[which],  # Selected PDB ID
    b = b,                          # B-factor vector
    trimmed = sel                   # Trimmed PDB object
  ))
}
```

```{r}
ids <- c("4AKE","1AKE","1E4Y") # Define the set of PDB IDs to analyze
plot_protein(ids, which = 1) # Plot the first protein’s B-factors
```

```{r}
plot_protein(ids, which = 2) 
# Plot the second protein’s B-factors
```

```{r}

plot_protein(ids, which = 3) # Plot the third protein’s B-factors
```

R uses chain = "A" and elety = "CA" as default arguments, so you only need to specify them if you want different values.
